The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-09-26 paper.tex
submission version

\begin{aug}
\author[B,C]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
\author[A,C]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}
}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e10}}
\address[C]{equal contribution}
\end{aug}

%\begin{review}{editor}
%\reviewer{\fnms{First} \snm{Editor}\address{\orgname{University or Company name}, \cny{Country}}}
%\reviewer{\fnms{Second} \snm{Editor}\address{\orgname{First University or Company name}, \cny{Country}
% and \orgname{Second University or Company name}, \cny{Country}}}
%\end{review}
%\begin{review}{solicited}
%\reviewer{\fnms{First} \snm{Solicited reviewer}\address{\orgname{University or Company name}, \cny{Country}}}
%\reviewer{\snm{anonymous reviewer}}
%\end{review}
%\begin{review}{open}
%\reviewer{\fnms{First} \snm{Open Reviewer}\address{\orgname{University or Company name}, \cny{Country}}}
%\end{review}

\begin{abstract}
The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
ANNO consists of two components:
ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
The integration is based on the three-ontology method with the General Formal Ontology as the top-level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically, .
ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
\end{abstract}

@@

ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures. In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains.
ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.

\begin{table}[b]
\centering
\caption{Overview of ANNO.}
\label{tab:publication}
\begin{tabulary}{\textwidth}{ll}
\toprule
%Property			&Value\\
%\midrule
Modules				&ANNOdc (domain-core) , ANNOds (domain-specific)\\
Top-Level Ontology	&General Formal Ontology~\citep{gfo}\\
Namespace			&\url{https://annosaxfdm.de/ontology/}\\
Terminology Server	&\url{https://ols.imise.uni-leipzig.de/ontologies/anno}\\
License				&\href{https://creativecommons.org/publicdomain/zero/1.0/}{Creative Commons Zero v1.0 Universal}\\
DOI (all versions)		&\href{https://doi.org/10.5281/zenodo.8301559}{10.5281/zenodo.8301559}\\
\bottomrule
\end{tabulary}
\end{table}

@@

An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of a bone's dimensions.
An anatomical line is a boundary (one-dimensional) of an anatomical surface.
Anatomical lines can connect or pass through anatomical entities (e.g., an edge between two or an angle between three anatomical entities).
The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes.% [Verweis zu Phenotype section].

@@

While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
This approach ensures intuitive and unimpeded data input and a valid end result.
The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.

In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
%
%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
%\subsections{General considerations}
%All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
%
%Anatomical Landmarks (Bone Parts) (Selection Criteria, scope and process, properties)
%Osteometric Landmarks (Anatomical Points)
%Measurements (Anatomical line)
%Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
%Phenotypes
%Development of the ontology is based on the following selection criteria:
%The selection criteria are as follows:
Given the overwhelming number of bone parts present in the cranium, the initial selection was narrowed down to a selection of representative and relevant parts.
Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
Those structures included in the definitions of others were also to be defined.
For bones that lie on the median sagittal plane (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
For osteometric landmarks, all those already established in the core literature are to be used.
Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
%In addition, it also had to be integrable into \aw{}.
The functions chosen were those with diagnostic value.
These were discriminant functions for sex determination and regression functions for body height and body weight estimation.

@@

%\subsection{}\label{s1.1}
\section{Conclusion and Future Work}
% "desirablility" add source e.g. https://fipat.library.dal.ca/wp-content/uploads/2021/08/FIPAT-TA2-Part-2.pdf

@@

%\begin{table*}
%\caption{} \label{t1}
%\begin{tabular}{lll}
%\hline
%&&\\
%&&\\
%\hline
%\end{tabular}
%\end{table*}

@@

%\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e4]{laura.penne@hs-mittweida.de}}
%\author[B]{\inits{A.}\fnms{Andrea} \snm{Ferencová}\ead[label=e7]{andrea.ferencova@hs-mittweida.de}}
%\author[B]{\inits{N.}\fnms{Niklas} \snm{de Sousa Norte}\ead[label=e6]{niklas.desousanorte@hs-mittweida.de}}
%\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e5]{fritzsc2@hs-mittweida.de}}
%\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e8]{andy.ludwig@hs-mittweida.de}}% ist jetzt beim Fraunhofer, möchte aber weiter so gelistet werden
%\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e9]{kreuzer@hs-mittweida.de}}
\end{minipage}%
\hfill%
\begin{minipage}{0.04\textwidth}\raggedleft
\includegraphics[width=\textwidth]{img/saxony.pdf}
\end{minipage}
\end{ack}

>>>>

\begin{aug}
\author[B,C]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
\author[A,C]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}
}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de}}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexandr.uciteli@imise.uni-leipzig.de}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e10}}
\address[C]{equal contribution}
\end{aug}

\begin{abstract}
The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
ANNO consists of two components:
ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
The integration is based on the three-ontology method with the General Formal Ontology as the top-level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically.
ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
\end{abstract}

@@

ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.
%ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.

\begin{table}[b]
\centering
\caption{Structure, properties and publication of ANNO.}
\label{tab:publication}
\begin{tabulary}{\textwidth}{ll}
\toprule
%Property			&Value\\
%\midrule
Modules				&ANNOdc (domain-core) , ANNOds (domain-specific)\\
Top-Level Ontology	&General Formal Ontology~\citep{gfo}\\
Namespace			&\url{https://annosaxfdm.de/ontology/}\\
Browser			&RickView~\citep{rickview} at \url{https://annosaxfdm.de/ontology/}\\
Repository		&\url{https://github.com/annosaxfdm/ontology}\\
Terminology Server	&\url{https://ols.imise.uni-leipzig.de/ontologies/anno}\\
License				&\href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International}\\
DOI (all versions)		&\href{https://doi.org/10.5281/zenodo.8301559}{10.5281/zenodo.8301559}\\
\bottomrule
\end{tabulary}
\end{table}

@@

An anatomical line is a conceptual, immaterial line on a bone passing between at least two measurement points that, in the form of distances, circumferences or angles is used to collect measurements representing aspects of a bone's dimensions.
An anatomical line is a boundary (one-dimensional) of an anatomical surface.
Anatomical lines can connect or pass through anatomical entities (e.g., an edge between two or an angle between three anatomical entities).
The length of the line or the angle degree can be measured and used in functions to infer individual phenotypes.% [Verweis zu Phenotype section].
For example, the anatomical line \anno{ZyDexterumZySinistrum} between the \emph{Zygion Dexterum} and the \emph{Zygion Sinistrum} is used in the sex determination function for the phenotype \anno{SexGilesElliot196319}~\citep{sexgileselliot1963}.

@@

While ANNOdc is created by the ontologists in consultation with the domain experts, ANNOds is developed by the domain experts themselves.
%The primary focus is on the design of ANNO's ontological architecture and the conception of the process for creating the content, as well as on the integration into the Applied University of Mittweida's in-house software AnthroWorks3D.
For this purpose they were provided with a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}, eliminating the requirement of having a background in RDF and ontologies.
The template is based on the structure of ANNOdc, so that the entered data is compliant with it: The ANNOds classes are subclasses of the ANNOdc classes (see \cref{tab:core}) and properties (see \cref{fig:core}) from ANNOdc are used.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
This approach ensures intuitive and unimpeded data input and a valid end result.
Additionally, ANNOds is validated using SHACL shapes\footnote{Contained in \texttt{dist/shacl.ttl} in \url{https://zenodo.org/record/8380382}.}, which requires metaclasses.
For example, all directly specified and transitive subclasses of \anno{Bone} are also explicitly individuals of the metaclass \anno{BoneClass} because of the limitations of SHACL.
In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
%
%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
%\subsections{General considerations}
%All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
%
%Anatomical Landmarks (Bone Parts) (Selection Criteria, scope and process, properties)
%Osteometric Landmarks (Anatomical Points)
%Measurements (Anatomical line)
%Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
%Phenotypes
%Development of the ontology is based on the following selection criteria:
%The selection criteria are as follows:
Given the overwhelming number of bone parts present in the cranium, the initial selection was narrowed down to a selection of representative and relevant parts.
Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
Those structures included in the definitions of others were also to be defined.
For bones that lie on the median sagittal plane (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
For osteometric landmarks, all those already established in the core literature are to be used.
Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
%In addition, it also had to be integrable into \aw{}.
The functions chosen were those with diagnostic value.
These were discriminant functions for sex determination and regression functions for body height and body weight estimation.

@@

\section{Conclusion and Future Work}
% "desirablility" add source e.g. https://fipat.library.dal.ca/wp-content/uploads/2021/08/FIPAT-TA2-Part-2.pdf

@@

@@

\end{minipage}%
\hfill%
\begin{minipage}{0.04\textwidth}\raggedleft
\includegraphics[width=\textwidth]{img/saxony.pdf}
\end{minipage}
\end{ack}
